### 2017-12-10 paper/jfp-submission/response.txt

We thank the reviewers for their detailed comments and suggestions.

## Questions about the user study

Reviewer 1
> I was not completely convinced by the metrics used for the statistical
> analysis; I would have found it more convincing to measure the speed
> at which an incorrect program may be fixed by a student sitting in front
> of a machine. That said, some statistical analysis is better than none,
> and I am greatful that the authors attempted some evaluation.

Reviewer 2
> 7. why did you do a user study as part of an exam, and not as an
> independent study? I don't think this is a good idea.

We agree that it would be preferable to have timing data in addition to
correctness data. We did attempt a study that would have collected
timing data, but unfortunately were not able to recruit enough
participants to draw any conclusions. We have updated the text in the
"Constuct" subsection of the "Threats to Validity" to mention lack of
timing data as a threat.

Reviewer 2
> 2. you compare only to OCAML in your student user study, which seems
> to be to be an easy target. Why not Sherrloc or Mycroft which at least
> get the location right more often? If OCAML much more often gives the
> wrong location, the message is not likely to be helpful. Indeed, I
> would expect you to only compare the systems for those cases that both
> NANOMALY and OCAML blame the same location.

Reviewer 2
> 3. does your tool help students to learn about the type system, or is
> it simply a better crutch to help them along. Did you investigate this
> at all?

These are valid concerns, we have updated the "Threats to Validity" to
mention them.

Reviewer 3
> The discussion of inter-rater reliability is strange. Why do we need
> to know if the exams are graded similarly, except to check that UCSD
> students are fairly graded? Also, why is IRR so low, given that
> these are questions with a correct answer? Given that there are only
> 60 exams in question, having a single evaluator with a clear rubric
> would eliminate both this potential source of inconsistency as well
> as the need to discuss IRR at all.

Our use of IRR is intended to account for potential bias in the
raters. One such form of bias (conscious or not) could be bias in favor
of NanoMaLy since it is new and interesting, and the raters (even if
they are not directly connected to the project) may want to see it
succeed. Due to the graphical nature of NanoMaLy's traces it is not easy
to hide whether a given answer was assisted by OCaml or NanoMaLy from
the raters. A further form of potential bias is inconsistency in
evaluating the student answers, which were long-form, English
explanations of the error/fix. There is a surprising amount of
subjectivity involved in judging, based on a couple sentences, whether
the student truly understood the error.

For these reasons we feel that our use of IRR is justified to lend
credibility to the results.

(We have not added this discussion to the paper as the IRR test is
really meant to be a preliminary "sanity-check", but would be happy to
if the reviewers feel it would be helpful.)

## Reviewer 1

> The tool is restricted to a small subset of OCaml, in particular it does
> not support side-effects. While many FP courses begin by focusing on
> pure programs only, there are some intringuing type errors that students
> face only when working with effectful programs, e.g. forgetting to apply
> a function to the unit argument. I speculate that it would be /not/ so
> easy to generalize the present work to deal with side effects, because
> of the need to maintain the state and report it to the user.

TODO

> Random-test generation would likely fail to produce counter-examples
> on slightly more complex programs such as red-black trees; for such
> programs, more advanced techniques such as constraint-based test
> generation are probably required. I was not able to quite understand
> whether the approach presented in the paper could be extended to
> accomodate constraint-based test generation.

TODO

> page 23, regarding section "witness exists", I got puzzled. The study
> there is on 50 programs. It was found that 2% of the programs admit a
> witness. I deduce this must correspond to one single program. But
> then, the next sentence tells that "slightly over half of these
> programs involved had some particular property, while the rest had
> some other property."  There semms to be some inconcistency there.

Indeed, the wording here was confusing. The percentages in this section
are presented as a fraction of *all* programs, not just the random
sample of failing programs. We have updated the text at the start of 5.3
to explain this.

## Reviewer 2

> 1. you do not address the issue of bias, the fact that locations
> towards the end of the program are more likely to be crashed at. This
> is similar to the left-to-right-bias of algorithm W, although in your
> case the left-to-rightness is in the trace, not the syntax tree (as
> addressed for that case in work by Hage and Heeren (VODCA 2008) and
> the work of Erwig and Chen).  Does your implementation allow general
> datatypes so that I can have case with many arms?

TODO

> 5. your approximation of functions to fun is rather strong, and it is
> unclear to me how well that works. I guess since all executions start
> from the beginning of the programs keeps everything rather concrete,
> so that you do not need to invent functions. But I am not sure, and I
> would like to see a thorough discussion. I am fine with novice
> programs being short and simple since that is quite likely to be the
> case when they resort to Nanomaly, but when you do that, it should be
> clear that you do.  What may be related to this is the idea to deal
> with Currying. What about functions like foldr of which the result is
> a type variable? I know that some static work on error diagnosis
> sometimes runs into a problem of some sort and since you work at the
> dynamic level this may not be the case for you. But do say and explain
> so.

TODO

> 1. explain how things would be if applied to other functional
> languages, in particular Haskell. What about laziness? What about
> overloaded numerals?  It seems to me that these are quite
> problematic. For example, the 0 in your fac program will be Num a =>
> a.

TODO

> 2. what is the limit to the size of programs you can reasonably
> handle?  What if we have a large program, is it possible to have a
> different starting point? So does a trace have to start at the start
> of the program? I guess this is important since if you start with a
> higher order function then you need to dynamically invent a function,
> and then your approximation might well be in the way. This is
> something I am really curious about.  Now you have 5% Non-parametric
> function types. Does that change in this situation?

TOOD

### 2017-12-10 paper/jfp-submission/evaluation.tex
start responding to reviews

\paragraph{4. Witnesses Assign Blame}
Finally, we present a simple heuristic that allows us to use witnesses
to \emph{automatically} assign blame for type errors.
%
We treat the values inside the stuck term as \emph{sources} of typing
constraints and the stuck term itself as a \emph{sink}, producing
a slice of the program that likely contains the error.
%
Using this heuristic, \toolname's witnesses are competitive with the
state-of-the-art localization tools \mycroft and \sherrloc.

@@

\subparagraph{Construct.}
%
We used the correctness of the student's explanation of, and fix for,
the type error as a proxy for her understanding, but it is possible
that other metrics would produce different results.

>>>>

\paragraph{4. Witnesses Assign Blame}
Finally, we present a simple heuristic that allows us to use witnesses
to \emph{automatically} assign blame for type errors.
%
We treat the values inside the stuck term as \emph{sources} of typing
constraints and the stuck term itself as a \emph{sink}, producing
a slice of the program that likely contains the error.
%
Using this heuristic, \toolname's witnesses are competitive with the
state-of-the-art localization tools \mycroft~\cite{Loncaric2016-uk}
and \sherrloc~\cite{Zhang2014-lv}.

@@

\subparagraph{Construct.}
%
We used the correctness of the student's explanation of, and fix for,
the type error as a proxy for her understanding, but it is possible that
other metrics would produce different results. One such metric that is
also surely relevant is time-to-completion, \ie a good error report
should \emph{quickly} guide the student to a fix. Unfortunately, the
in-class exam setting of our study did not admit the collection of
timing data.

Furthermore, one might object to our selection of \ocaml as
the baseline comparison rather than \sherrloc or \mycroft, which also
claim to produce more accurate error reports. This is indeed a
limitation of our study, but we note that \sherrloc blames the same
expression as \ocaml in both @wwhile@ and @append@, and in @digitsOfInt@
both blame the wrong function.

Finally, one might point out that our study investigates the use of
\toolname as a \emph{debugging} aid rather than as a \emph{teaching}
aid. That is, it may be that \toolname helps students solve their
immediate problem, but does not help them build a lasting understanding
of the type system. We have not attempted a longitudinal study of the
long-term impact of using \toolname, but we agree that it would be an
interesting future direction.

### 2017-12-10 paper/jfp-submission/safe.tex
start responding to reviews

To answer this question, we investigated the 732 \ucsdbench programs for
which we failed to find a witness.
%
We used a combination of automatic and manual coding to categorize these
programs into four classes.
%
The first class is easily detected by \toolname itself, and thus admits
a precise count.
%
This left us with 504 programs that required manual coding; we selected
a random sample of 50 programs to investigate, and will report results
based on that sample.
%
Figure~\ref{fig:no-witness} summarizes the results of our investigation ---
we note the classes that were based on the random sample with a ``*''.

@@

\paragraph{Ad-Hoc Polymorphism}
%
We found that for 5\% programs \toolname got stuck when it tried to
compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
For example, consider the \hbox{@n <= 0@} test in our @fac@ example.
%
The @<=@ operator is polymorphic, but in this case we can make progress because
the literal @0@ is not.
%
Suppose, however, we parameterized @fac@ by a lower bound, \eg
%
\begin{code}
  let rec fac n m =
    if n <= m then
      true
    else
      n * fac (n - 1) m
\end{code}
%
When given @fac@, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@n < m@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |n|
and |m| such that |n <= m| is false.

@@

\paragraph{Non-Parametric Function Type *}
%
5\% of programs lack a witness in our semantics due to our
non-parametric $\tfun$ type for functions.
%
Recall that our goal is to expose the runtime errors that would have
been prevented by the type systems.
%
At runtime, it is always safe to call a function, thus we give functions
a simple type $\tfun$ that says they may be applied, but says nothing
about their inputs or outputs.
%
But consider the following @clone@ function, which is supposed to
produce a list containing @n@ copies of the input @x@.
%
% Seven programs violated the \emph{occurs check} with cyclic typing
% constraints like @'a = 'a list@, for example the following @clone@
% function.
%
\begin{code}
  let rec clone x n =
    if n > 0 then
      clone [x] (n - 1)
    else
      []
\end{code}
%
Unfortunately, the student instead constructs an @n@-level nested list
containing a single @x@.
%
The \ocaml compiler rejects this program because the recursive call to
@clone@ induces a cyclic typing constraint @'a = 'a list@, capturing the
fact that each call increases the nesting of the list.
%
\toolname fails to catch this because we do not track the types of the
inputs to @clone@.

@@

\paragraph{Dead Code and ``Safe'' Function Calls *}
%
4\% of programs contained type errors that were unreachable, either
because they were dead code, or because the student called the function
with inputs that could not trigger the error.

@@

\paragraph{Witness Exists *}
%
We found that only 2\% of programs admit a witness that \toolname
was unable to discover.
%
Slightly over half involved synthesizing a \emph{pair} of
specially-crafted inputs that would result in the function returning
values of incompatible types.
%
The rest required synthesizing an input that would trigger a
particular path through the program, and would likely have been caught
by symbolic execution.

>>>>

To answer this question, we investigated the 732 \ucsdbench programs for
which we failed to find a witness.
%
We used a combination of automatic and manual coding to categorize these
programs into four classes.
%
The first class is easily detected by \toolname itself, and thus admits
a precise count.
%
This left us with 504 programs that required manual coding; we selected
a random sample of 50 programs to investigate, and will report results
based on that sample.
%
Figure~\ref{fig:no-witness} summarizes the results of our investigation ---
we note the classes that were based on the random sample with a ``*''.
%
Note that the percentages referenced in Figure~\ref{fig:no-witness} (and
in the sequel) are with respect to the total number of programs in the
\ucsdbench dataset, not only those were \toolname failed to find a
witness.

@@

\paragraph{Ad-Hoc Polymorphism}
%
We found that for 5\% of all programs \toolname got stuck when it tried to
compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
For example, consider the \hbox{@n <= 0@} test in our @fac@ example.
%
The @<=@ operator is polymorphic, but in this case we can make progress because
the literal @0@ is not.
%
Suppose, however, we parameterized @fac@ by a lower bound, \eg
%
\begin{code}
  let rec fac n m =
    if n <= m then
      true
    else
      n * fac (n - 1) m
\end{code}
%
When given @fac@, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@n < m@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |n|
and |m| such that |n <= m| is false.

@@

\paragraph{Non-Parametric Function Type *}
%
5\% of all programs lack a witness in our semantics due to our
non-parametric $\tfun$ type for functions.
%
Recall that our goal is to expose the runtime errors that would have
been prevented by the type systems.
%
At runtime, it is always safe to call a function, thus we give functions
a simple type $\tfun$ that says they may be applied, but says nothing
about their inputs or outputs.
%
But consider the following @clone@ function, which is supposed to
produce a list containing @n@ copies of the input @x@.
%
% Seven programs violated the \emph{occurs check} with cyclic typing
% constraints like @'a = 'a list@, for example the following @clone@
% function.
%
\begin{code}
  let rec clone x n =
    if n > 0 then
      clone [x] (n - 1)
    else
      []
\end{code}
%
Unfortunately, the student instead constructs an @n@-level nested list
containing a single @x@.
%
The \ocaml compiler rejects this program because the recursive call to
@clone@ induces a cyclic typing constraint @'a = 'a list@, capturing the
fact that each call increases the nesting of the list.
%
\toolname fails to catch this because we do not track the types of the
inputs to @clone@.

@@

\paragraph{Dead Code and ``Safe'' Function Calls *}
%
4\% of all programs contained type errors that were unreachable, either
because they were dead code, or because the student called the function
with inputs that could not trigger the error.

@@

\paragraph{Witness Exists *}
%
We found that only 2\% of all programs admit a witness that \toolname
was unable to discover.
%
Slightly over half involved synthesizing a \emph{pair} of
specially-crafted inputs that would result in the function returning
values of incompatible types.
%
The rest required synthesizing an input that would trigger a
particular path through the program, and would likely have been caught
by symbolic execution.
